Multidimensional Persistence Module Classification via Lattice-Theoretic Convolutions

### 2020-10-10 paper/paper.tex
add discussion

\author{%
  Hans Riess  \\
  Department of Electrical and Systems Engineering \\
  Universiyt of Pennsylvania\\
  Philadelphia, PA 19104 \\
  \texttt{hmr@seas.upenn.edu} \\
  % examples of more authors
  \And
  Jakob Hansen \\
  Department of Mathematics \\
  Ohio State University \\
  Columbus, OH 43210 \\
  \texttt{hansen.612@osu.edu}
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

@@

\begin{abstract}
 We consider the use of lattice-based convolutional neural network layers as a 
 tool for the analysis of features arising from multiparameter persistence
 modules. \textbf{CHANGE THIS: something something the conclusion about what to do}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

Multidimensional persistence suffers two hindrances to its usefulness in
machine learning. First, software for computing multidimensional persistence is
scarce. To the authors' knowledge, RIVET is the only available software for
computing multidimensional persistence~\cite{lesnick_interactive_2015}; RIVET
specializes to $2$-dimensional persistence and is focused on interactive
visualization rather than machine-interpretable output.
% provides a user-interface for
% visualizing multi-filtrations as well as a pipline for computing the Hilbert
% function and the multi-graded Betti numbers in degrees $0$, $1$, and $2$.
Second, %for all the attention multidimensional persistence has received,
there has been no activity, as far as the authors are aware, studying the extraction
of suitable features for multidimensional persistent homology.

@@

Due to space constraints, we can offer only a brief overview of multiparameter
persistent homology. For a primer on persistent homology,
see~\cite{ghrist_barcodes:_2008, carlsson_topology_2009}; for multiparameter
persistent homology, see~\cite{}. An introduction to lattices may be found
in~\cite{}.

@@

To this end, we construct a lattice convolution-based neural network layer
suitable for use with features originating from multidimensional persistence
modules. To the authors' knowledge, such an architecture has not previously been
described, although a special case (where the underlying lattice is a power set)
has been implemented in~\cite{wendler_powerset_2019}. We specialize the convolutions described in
Section~\ref{sec:latticeconv} to the particular case of regular finite
sublattices of $\R^2$. These may be represented (up to isomorphism)as $L = [m]
\times [n]$, where $[n]$ is the ordered set $\{0,1,\dots,n\}$. The meet and join operations are easily computed
elementwise:
\[(r,t) \wedge (r',t') = (\min(r,r'),\min(t,t'));\quad (r,t)\vee (r',t') =
  (\max(r,r'),\max(t,t')).\]

@@

\section{Experiments}
We use a small portion of the Princeton ModelNet
dataset~\cite{zhirong_wu_3d_2015} as a source of finite
metric spaces. This dataset consists of hundreds of 3-dimensional CAD models
representing objects from 40 classes. We select two of the classes and sample
points from the 3D models to produce finite metric spaces embedded in $\R^3$.
We then compute the corresponding multidimensional persistence modules, from
which we produce features used as an input to a convolutional neural net
classifier.

The pipline thus begins with a 3D model, which is sampled to produce a point
cloud in $\R^3$. This point cloud then produces a bifiltered simplicial complex,
whose persistent homology we calculate using
RIVET~\cite{lesnick_interactive_2015}, sampled at a discrete grid of $40 \times
40$ points, producing lattice-indexed signals given by the Hilbert function and
the multi-graded Betti numbers. These are then passed to the classifier, which
produces a class prediction.

@@

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by three fully connected layers.

\section{Discussion}
Discuss.

>>>>

\author{%
  Hans Riess  \\
  Department of Electrical and Systems Engineering \\
  University of Pennsylvania\\
  Philadelphia, PA 19104 \\
  \texttt{hmr@seas.upenn.edu} \\
  % examples of more authors
  \And
  Jakob Hansen \\
  Department of Mathematics \\
  Ohio State University \\
  Columbus, OH 43210 \\
  \texttt{hansen.612@osu.edu}
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

@@

\begin{abstract}
 Multiparameter persistent homology has been largely neglected as an input to
 machine learning algorithms.
 We consider the use of lattice-based convolutional neural network layers as a 
 tool for the analysis of features arising from multiparameter persistence
 modules. We find that these show promise as an alternative to convolutions for
 the classification of multidimensional persistence modules.
\end{abstract}

\section{Introduction}

@@

Multidimensional persistence suffers two hindrances to its usefulness in
machine learning. First, software for computing multidimensional persistence is
scarce. To the authors' knowledge, RIVET is the only available software for
computing multidimensional persistence~\cite{lesnick_interactive_2015}; RIVET
specializes to $2$-dimensional persistence and is focused on interactive
visualization rather than machine-interpretable output.
% provides a user-interface for
% visualizing multi-filtrations as well as a pipline for computing the Hilbert
% function and the multi-graded Betti numbers in degrees $0$, $1$, and $2$.
Second, %for all the attention multidimensional persistence has received,
there has been no activity, as far as the authors are aware, studying the extraction
of suitable features for multidimensional persistent homology.
%Carlsson has a little bit in a recent paper...

@@

Due to space constraints, we can offer only a brief overview of multiparameter
persistent homology. For a primer on persistent homology,
see~\cite{ghrist_barcodes:_2008, carlsson_topology_2009}; for multiparameter
persistent homology, see~\cite{lesnick_interactive_2015}. An introduction to
lattices may be found in~\cite{}.

@@

To this end, we construct a lattice convolution-based neural network layer
suitable for use with features originating from multidimensional persistence
modules. To the authors' knowledge, such an architecture has not previously been
described, although a special case (where the underlying lattice is a power set)
has been implemented in~\cite{wendler_powerset_2019}. We specialize the convolutions described in
Section~\ref{sec:latticeconv} to the particular case of regular finite
sublattices of $\R^2$. These may be represented (up to isomorphism) as $L = [m]
\times [n]$, where $[n]$ is the ordered set $\{0,1,\dots,n\}$. The meet and join operations are easily computed
elementwise:
\[(r,t) \wedge (r',t') = (\min(r,r'),\min(t,t'));\quad (r,t)\vee (r',t') =
  (\max(r,r'),\max(t,t')).\]

@@

\section{Experiments}
\begin{figure}
  \caption{Comparison training curves for the lattice neural network and
    standard convolutional neural network.}\label{fig:comparison}
\end{figure}

We use a small portion of the Princeton ModelNet
dataset~\cite{zhirong_wu_3d_2015} as a source of finite
metric spaces. This dataset consists of hundreds of 3-dimensional CAD models
representing objects from 40 classes. We select two of the classes and sample
points from the 3D models to produce finite metric spaces embedded in $\R^3$.
We then compute the corresponding multidimensional persistence modules, from
which we produce features used as an input to a convolutional neural net
classifier.

The pipline thus begins with a 3D model, which is sampled to produce a point
cloud in $\R^3$. This point cloud then produces a bifiltered simplicial complex,
whose persistent homology we calculate using
RIVET~\cite{lesnick_interactive_2015}, sampled at a discrete grid of $40 \times
40$ points, producing lattice-indexed signals given by the Hilbert function and
the multi-graded Betti numbers. These are then passed to the classifier, which
produces a class prediction, in this case either ``sofa'' or ``monitor.''

@@

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by three fully connected layers.
The lattice-based convolution layers are of the form $\text{MeetConv}(x) +
\text{JoinConv}(x)$. All convolution kernels have dimension $4 \times 4$, hidden
convolution layers have 16 features, and the final convolution has 8 features.
The fully connected layers have 32 features at each inner layer.
The networks are trained using the Adam gradient algorithm with learning rate
$0.001$ for 50 epochs. We reserve 10\% of the data for testing. Results are
shown in Figure~\ref{fig:comparison}.

The lattice-based convolution and classical convolution networks perform
similarly on the training set. However, the lattice convolution network's
predictions generalize more accurately to the testing set.
%% more discussion of what the data actually looks like

\section{Discussion}
Our proposed featurization for persistence modules is rather naive, but well
adapted to the use of lattice convolutions as a data processing method. The
lattice convolutional neural network shows promise as a method for classifying
features arising from a multiparameter persistence module. The algebraic
perspective on partially ordered sets, exemplified by lattices, may also offer
approaches to featurizing more complex invariants of persistence modules. In
particular, the incidence algebra may offer a natural way to represent the rank
invariant~\cite{carlsson_theory_2009} in a way amenable to convolution-like
operations.

We hope with these brief experiments to inspire further work on featurizing
multidimensional persistence for use in machine learning algorithms.
% anything else?
